Technology for application family creation based on domain analysis

### 2016-04-10 2016-domainAnalysis/domainAnalysis.tex
Added reference to Koznov, also some fixes to make this article more readable

\begin{abstract}
The theme of code reuse in software development is hot-button issue. Sometimes it is hard to find out what exactly we need to reuse in isolation of context. However, there is an opportunity to narrow the context problem, if applications in one given domain are considered. Hence, the problem of domain analysis arises. On the other hand, there is metaCASE-techonology for application code generation with diagrams, which describe the given application. The main objective of this article is to present the technology for application family creation, which connects the metaCASE-techonology and domain analysis. We propose to use feature diagrams to describe variability in a domain and then create domain-specific visual language that allows to connect and configure existing feature implementations thus producing an application. This technology supposed to be especially useful for software product lines.
\end{abstract}

@@

\section{Introduction}
The term ``reuse'' in software engineering closely associated with context. Reuse objects can be programs, parts of programs, specifications, requirements, architectures, test plans, etc. Reuse of one object leads to reuse of  another object. This means, there is a need to reuse something more than just code, i.e. there is a call for increasing the abstraction level. It is commonly supposed that reuse, as some kind of activity, can be divided into groups according to what should be reused: components, process for gaining the product, the technology or knowledge. At all accounts any reuse object cannot be discussed without environment, where the given object exists. Hence, the context problem is still remains.

@@

Gathering information about the domain is the crucial step in the whole process of software development. Nowadays, applications in one domain are often designed independently; this approach leads to increase of development time and cost. Usually such applications have similar functionality, so the reuse problem moves to the forefront in an attempt to speed up the development and to decrease the cost for systems in one domain. The reuse process in one domain supposes the necessity of the domain analysis activity. At present domain analysis in software life cycle is performed in informal way. There are some domain analysis tools, but such tools are not integrated with development tools. As the result of the domain analysis activity some diagrams just are put up on the board, and do not take part in following process of software design. The risk of incorrect understanding of domain-dependent knowledge increases. Therefore, many peculiarities of the domain may be missed in development process because of the factor of human error. This fact may lead to development of the product which do not satisfies requirements at all. Hence, there emerged a need for a tool, in which domain analysis activity would play a vital role in software development process, i.e. based on this activity would be possible to generate some design model, so developers and other process actors could rely on this model. At the present day there is no tool that could allow to solve this problem.

@@

Main idea of domain-specific modeling is to use a number of visual languages in one tool to develop a complete system. Every language can provide a different point of view on a system. We propose to exploit this idea to automatically produce useful artefacts from the results of domain analysis thus seamlessly integrating this phase into development process. For that, we will use specific visual language to perform domain analysis and to build domain model, language simple enough to be useful to analysts and domain experts who do not necessarily possess programming skills. Then, using this domain model, we will generate actual domain-specific language that will allow to configure various existing pre-built components and integrate them to generate a working application. As we will see, this language will also typically be very simple so that non-programmers can use it. The only real coding in the proposed approach occurs when creating components from which applications will be built, but for product lines these components will already exist anyway, as they will in a case when a team develops many applications in one domain for some time. Not all steps in proposed approach are fully automatic, as a visual language needs tailoring after generation from domain model --- we still need to manually specify shapes of its elements (to be familiar for domain experts) and configure properties which depend on existing components and can not be derived from domain model. It is also possible that generated application will need tailoring by hand, but generation can significantly lower the effort needed to create application.

@@

\section{Domain analysis approaches}
\label{chapter:approaches}
There is no any clear and long-standing definition of the term ``domain analysis''. Almost all papers, in which this term is considered, go back to 80s-90s of the twentieth century. It was then that scientists, taking into account rapidly growing technologies, were thinking about global reuse. Always projects are developing for concrete user needs, so then the term ``domain'' took the definition. Domain is the field of expertise, problems in which the software intend to solve. According to Rugaber~\cite{rugaber1994domain}, the domain is described in terms of glossary, some assumptions, architecture approach and literature.

Then the question arise, how we need to analyze the domain for acquiring the necessary information. At present, the information gathering into knowledge bases is understood under the term ``domain analysis''. Although, Prieto-Diaz~\cite{prieto1988domain} confirms that domain analysis is an activity, which take place before system analysis and its output is used for system analysis to the same degree as system analysis’s output is used for system design.  There are other definitions of the term ``domain analysis''. Ferre~\cite{ferre1999evaluation} has presented definitions, such as: (1) the process of identification, organization and presenting the relevant information of a given domain, (2) the process, in which the customer’s knowledge are identified, concretized and systemized. The relevant information of the domain should be presented in objective, readily available way, such way is called ``domain model''. Mernik~\cite{mernik2005} specifies, that the domain model include not only glossary, but also must describe commonalities and variabilities of terms. Such model should precisely set bounds of the domain, i.e. clear and exact characterize a range of questions, which are considered in the domain. Term variabilities allow to define exactly, what information must be specified in concrete system implementation. Term commonalities are used for defining a set of shared operations between different applications. Implementing commonalities and adding the gained model  with information, which can be specified in instance of the concrete system, a set of different system can be obtained based only one common model. In such manner, based only one domain model can be implemented the set of different systems in given domain. Taking into account definitions above mentioned, we can conclude that domain analysis is the activity forward system analysis, which goal is providing the domain model.

As stated above, at present in many software companies the term ``domain analysis'' is understood as information gathering into some knowledge bases, but it is obviously that there are disadvantages of this approach. It may lead to incomplete glossary, absence of agreements about understanding some terms in the domain, so any misunderstanding of domain can result an improper product. Therefore, several dozens of years ago were introduced some formal approaches for domain analysis. Here will be mentioned some of them. It is worth noting that main objective any domain analysis approach is to gain the domain model.

Despite different understanding the term ``domain analysis'' Arango~\cite{arango1994domain} showed that all formal domain analysis methods follow the general process for obtain the domain model. This process includes next stages: (1) domain characterization, (2) data collection, (3) data analysis, (4) classification and finally (5) evaluation of domain model. There are following domain analysis approaches: 1) DARE (Domain Analysis and Reuse Environment)~\cite{frakes1998dare}. The crucial idea of this method is creating domain book. This book include the universal architecture and library of reusable components. 2) DSSA (Domain-Specific Software Architectures)~\cite{taylor1995software}. Given approach allows to create a domain glossary with the aid of use case analysis. 3) ODE (Ontology-based Domain Engineering)~\cite{falbo2002ontological}. This approach connects the ontology idea with object-oriented approach. Ontology includes  terms and their connections, definitions, properties and constraints. Library of objects is built based on mapping ontology with object-oriented entities. 4) FODA (Feature-Oriented Domain Analysis)~\cite{kang1990feature}. This method has get popularity among scientists in the research area because of its simplicity for non-programmers. The main idea of this approach is creating feature model. This model describes functionality, which the future product should possess. Such model must note, what features are compulsory for implement in any instance of application in a given domain, what features must be implemented but there is some alternative between them, and present features, which may be implemented but not compulsory. This model can be easily built by expert in the domain.

Concerning product line creating with the aid of using domain model, Estublier~\cite{estublier2005reuse} presented approach, which is based on some aspects and requirements. These entities were proposed by authors. Such approach based on MDE methodology. Domain model is considered as  metamodel, which is described on MOF or UML. There is an interpreter, which translates each term in metamodel into Java-class, and concrete models --- into instances of these classes. Domain model is accompanied with feature model, which include some external behavior of the system. Authors use aspect-oriented techniques for feature implementing and following their  mapping with terms in domain model. Consequently, there is a close interaction between domain modeling and feature modeling. It seems that such approach is a bit complicated for non-programmers. In addition, there is no any industrial use of this method, but it is worth noting that authors describe appliance in this article~\cite{estublier2003approach}.

@@

\section{Evaluation}
\label{chapter:evaluation}
For efficiency demonstrating of proposed above approach there was implemented a model application for  remote control of various robot models --- ``Joystick''. The main substantiation for implementing such application is that controlling different robot models requires different control elements. For example, one model can be controlled with only two pads, but another --- with one pad and two buttons. Such application was implemented in C\# for Windows Phone platform. Screenshots of this simple application are presented on Fig.~\ref{image:joystick}.

\begin{figure*}[t]
	\centering
	\includegraphics[width=0.4\textwidth]{joystick1.png}
	\includegraphics[width=0.4\textwidth]{joystick2.png}
	\caption{Screenshots of ``Joystick'' application.}
	\label{image:joystick}
\end{figure*}

As mentioned above, it was used QReal as DSM tool. There was implemented a visual language for describing feature models. Appropriate feature model for application family ``Joystick'' is proposed on Fig.~\ref{image:joystickFeatureModel}.

@@

Based on this feature model there was generated metamodel for future visual language, which is required for building different models for different configurations. Generated metamodel is presented on Fig.~\ref{image:joystickMetamodel}. As it can be seen, metamodel is very simple. At this stage we can propose that entities, such as ``buttons'' and ``pads'' may have a property ``Quantity''. In addition, we can set forms for these entities, which will be shown in visual language.

@@

Then with resources of QReal tool was generated a visual language. Example of generated visual language is demonstrated on Fig.~\ref{image:joystickDsl}. It can be seen that in visual language editor we can specify property ``Quantity'', explicitly pointing out concrete number of pads.

@@

To tell the truth, example is quite simple for demonstrating extensive possibilities proposed above approach. At present there is no rigorous evaluation of the offered process. It is worth noting that it is just an idea, there is no yet cohesive and consistent technology for creating application family based on domain analysis. Therefore, this work requires more detailed explorations.

\section{Conclusion}
\label{chapter:conclusion}
The problem of not using domain analysis result for further generating some entities for software development process was stated. There were considered some formal domain analysis approaches and we concluded that creation of feature diagrams is the most elegant decision for domain analysis that can be conducted by domain expert, i.e. non-programmer, may be in collaboration with system analysts. Moreover, there was discussed one of the possible solutions, which is presented by Estublier, we specify some problems of such method. We suggested our own approach for creation application family in one domain based on domain analysis. Also, there was some evaluation of this approach, where we pointed out that this example remains many questions because of its simplicity. Finally, this research requires more detailed explorations.

>>>>

\begin{abstract}
The theme of code reuse in software development is still important. Sometimes it is hard to find out what exactly we need to reuse in isolation of context. However, there is an opportunity to narrow the context problem, if applications in one given domain are considered. Hence, the problem of domain analysis arises. On the other hand, there is metaCASE-techonology that allows to generate code of an application using diagrams. The main objective of this article is to present the technology for application family creation which connects the metaCASE-techonology and domain analysis. We propose to use feature diagrams to describe variability in a domain and then create domain-specific visual language that allows to connect and configure existing feature implementations thus producing an application. This technology supposed to be especially useful for software product lines.
\end{abstract}

@@

\section{Introduction}
The term ``reuse'' in software engineering is closely associated with context. Reuse objects can be programs, parts of programs, specifications, requirements, architectures, test plans, etc. Reuse of one object leads to reuse of another object. This means, there is a need to reuse something more than just code, i.e. there is a call for increasing the abstraction level. It is commonly supposed that reuse, as some kind of activity, can be divided into groups according to what should be reused: components, process for gaining the product, technology or knowledge. At all accounts any reuse object cannot be discussed without environment, where the given object exists. Hence, the context problem still remains.

@@

Gathering information about the domain is the crucial step in the whole process of software development. Nowadays applications in one domain are often designed independently; this approach leads to increase of development time and cost. Usually such applications have similar functionality, so the reuse problem moves to the forefront in an attempt to speed up the development and to decrease the cost for systems in one domain. The reuse process in one domain supposes the necessity of the domain analysis activity. At present domain analysis in software life cycle is performed in informal way. There are some domain analysis tools, but such tools are not integrated with development tools. As the result of the domain analysis activity some diagrams just are put up on the board, and do not take part in following process of software design. The risk of incorrect understanding of domain-dependent knowledge increases. Therefore, many peculiarities of the domain may be missed in development process because of the factor of human error. This fact may lead to development of the product which does not satisfy requirements at all. Hence, there emerged a need for a tool in which domain analysis activity would play a vital role in software development process, i.e. based on this activity would be possible to generate some design model, so developers and other process actors could rely on this model. At the present day there is no tool that could allow to solve this problem.

@@

Main idea of domain-specific modeling is to use a number of visual languages in one tool to develop a complete system. Every language can provide a different point of view on a system. We propose to exploit this idea to automatically produce useful artefacts from the results of domain analysis thus seamlessly integrating this phase into development process (such as~\cite{koznov2011process}). For that, we will use specific visual language to perform domain analysis and to build domain model, language simple enough to be useful to analysts and domain experts who do not necessarily possess programming skills. Then, using this domain model, we will generate actual domain-specific language that will allow to configure various existing pre-built components and integrate them to generate a working application. As we will see, this language will also typically be very simple so that non-programmers can use it. The only real coding in the proposed approach occurs when creating components from which applications will be built, but for product lines these components will already exist anyway, as they will in a case when a team develops many applications in one domain for some time. Not all steps in proposed approach are fully automatic, as a visual language needs tailoring after generation from domain model --- we still need to manually specify shapes of its elements (to be familiar for domain experts) and configure properties which depend on existing components and can not be derived from domain model. It is also possible that generated application will need tailoring by hand, but generation can significantly lower the effort needed to create application.

@@

\section{Domain analysis approaches}
\label{chapter:approaches}
There is no any clear and long-standing definition of the term ``domain analysis''. Almost all papers, in which this term is considered, go back to 80s-90s of the twentieth century. It was then that scientists, taking into account rapidly growing technologies, were thinking about global reuse. Always projects are developing for concrete user needs, so then the term ``domain'' took the definition. Domain is the field of expertise, problems in which the software intends to solve. According to Rugaber~\cite{rugaber1994domain}, the domain is described in terms of glossary, some assumptions, architecture approach and literature.

Then the question arise, how we need to analyze the domain for acquiring the necessary information. At present, the information gathering into knowledge bases is understood under the term ``domain analysis''. Although, Prieto-Diaz~\cite{prieto1988domain} confirms that domain analysis is an activity, which take place before system analysis and its output is used for system analysis to the same degree as system analysis’s output is used for system design.  There are other definitions of the term ``domain analysis''. Ferre~\cite{ferre1999evaluation} has presented definitions, such as: (1) the process of identification, organization and presenting the relevant information of a given domain, (2) the process, in which the customer’s knowledge are identified, concretized and systemized. The relevant information of the domain should be presented in objective, readily available way, such way is called ``domain model''. Mernik~\cite{mernik2005} specifies that the domain model includes not only glossary, but also must describe commonalities and variabilities of terms. Such model should precisely set bounds of the domain, i.e. clear and exact characterize a range of questions, which are considered in the domain. Term variabilities allow to define exactly, what information must be specified in concrete system implementation. Term commonalities are used for defining a set of shared operations between different applications. Implementing commonalities and adding the gained model  with information, which can be specified in instance of the concrete system, a set of different systems can be obtained based on one common model. In such manner, based on one domain model, the set of different systems in given domain can be implemented. Taking into account definitions above mentioned, we can conclude that domain analysis is the activity of forward system analysis, which goal is to provide the domain model.

As stated above, at present in many software companies the term ``domain analysis'' is understood as information gathering into some knowledge bases, but it is obvious that there are disadvantages of this approach. It may lead to incomplete glossary, absence of agreements about understanding some terms in the domain, so any misunderstanding of domain can result in an improper product. Therefore, several dozens of years ago were introduced some formal approaches for domain analysis. Here will be mentioned some of them. It is worth noting that main objective any domain analysis approach is to gain the domain model.

Despite different understanding of the term ``domain analysis'', Arango~\cite{arango1994domain} showed that all formal domain analysis methods follow the general process for obtaining the domain model. This process includes next stages: (1) domain characterization, (2) data collection, (3) data analysis, (4) classification and finally (5) evaluation of domain model. There are following domain analysis approaches: 1) DARE (Domain Analysis and Reuse Environment)~\cite{frakes1998dare}. The crucial idea of this method is to create the domain book, that will include the universal architecture and library of reusable components. 2) DSSA (Domain-Specific Software Architectures)~\cite{taylor1995software}. Given approach allows to create a domain glossary with the aid of use case analysis. 3) ODE (Ontology-based Domain Engineering)~\cite{falbo2002ontological}. This approach connects the ontology idea with object-oriented approach. Ontology includes  terms and their connections, definitions, properties and constraints. Library of objects is built based on mapping ontology with object-oriented entities. 4) FODA (Feature-Oriented Domain Analysis)~\cite{kang1990feature}. This method has get popularity among scientists in the research area because of its simplicity for non-programmers. The main idea of this approach is creating feature model. This model describes functionality, which the future product should possess. Such model must note what features are compulsory for implement in any instance of application in a given domain, what features must be implemented but there is some alternative between them, and present features, which may be implemented but not compulsory. This model can be easily built by expert in the domain.

Concerning product line creating with the aid of using domain model, Estublier~\cite{estublier2005reuse} presented approach which is based on some aspects and requirements. These entities were proposed by authors. Such approach based on MDE methodology. Domain model is considered as  metamodel, which is described on MOF or UML. There is an interpreter, which translates each term in metamodel into Java class, and concrete models --- into instances of these classes. Domain model is accompanied with feature model, which include some external behavior of the system. Authors use aspect-oriented techniques for feature implementing and following their mapping with terms in domain model. Consequently, there is a close interaction between domain modeling and feature modeling. It seems that such approach is a bit complicated for non-programmers. In addition, there is no any industrial use of this method, but it is worth noting that authors describe appliance in this article~\cite{estublier2003approach}.

@@

\section{Evaluation}
\label{chapter:evaluation}
For demonstration of the efficiency of proposed above approach there was implemented a model application for remote control of various robot models --- ``Joystick''. The main substantiation for implementing such application is that controlling different robot models requires different control elements. For example, one model can be controlled with only two pads, but another --- with one pad and two buttons. Such application was implemented in C\# for Windows Phone platform. Screenshots of this simple application are presented on Fig.~\ref{image:joystick}.

\begin{figure*}[t]
	\centering
	\includegraphics[width=0.35\textwidth]{joystick1.png}
	\includegraphics[width=0.35\textwidth]{joystick2.png}
	\caption{Screenshots of ``Joystick'' application.}
	\label{image:joystick}
\end{figure*}

As mentioned above, it was used QReal as DSM tool. A visual language was implemented there for describing feature models. Appropriate feature model for ``Joystick'' application family is proposed on Fig.~\ref{image:joystickFeatureModel}.

@@

Based on this feature model a metamodel for future visual language was generated, which is required for building different models for different configurations. Generated metamodel is presented on Fig.~\ref{image:joystickMetamodel}. As it can be seen, metamodel is very simple. At this stage we can propose that entities, such as ``buttons'' and ``pads'', may have a property ``Quantity''. In addition, we can specify images for these entities, which will be shown in visual language.

@@

Then with help of QReal tool a visual language was generated. Example of generated visual language is demonstrated on Fig.~\ref{image:joystickDsl}. It can be seen that in visual language editor we can specify property ``Quantity'', explicitly pointing out concrete number of pads.

@@

To tell the truth, example is quite simple for demonstrating extensive possibilities of the approach proposed above. At present there is no rigorous evaluation of the proposed process. Also, cohesive and consistent technology for creating application family based on domain analysis is not implemented yet, here we have described a concept-proof prototype. Therefore, this work requires more detailed explorations.

\section{Conclusion}
\label{chapter:conclusion}
The problem of not using domain analysis result for further generation of some entities for software development process was stated. There were considered some formal domain analysis approaches and we concluded that creation of feature diagrams is the most elegant decision for domain analysis that can be conducted by domain expert, i.e. non-programmer, maybe in collaboration with system analysts. Moreover, there was discussed one of the possible solutions, which is presented by Estublier, we specify some problems of such method. We suggested our own approach for creation application family in one domain based on domain analysis. Also, there was some evaluation of this approach, where we pointed out that this example remains many questions because of its simplicity.
